The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-02-12 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further text revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2462 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 00f139ebb2799a8a337f5ea1a28509e978aaea20

\section{Introduction}\label{sec:introduction}
In many situations fitting one global model to a given data set can be very challenging, 
especially if the data contains lots of different features with a strong variation 
or complex interactions. Therefore, separating the data into more homogeneous subgroups 
based on a set of covariates first can simplify the task and fitting a local model to 
each of the resulting subgroups often leads to better results. This separation can be 
done by applying a tree algorithm. While all of the many developed algorithms 
follow the general idea of splitting the data such that some objective function is optimized, 
they differ in their specific approaches to selecting a split variable and the corresponding 
split point. 
Some of the first tree algorithms (e.g., AID, \citealp{Morgan+Sonquist:1963}; 
CART, \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
the best split point and split variable in one step by directly comparing all possible split
points in all possible split variables. However, it has been shown that this is not only
computationally expensive but also biased towards split variables with many possible split
points \citep{Doyle:1973, Kim+Loh:2001}. 
Therefore, selecting a split variable in a first step and then searching for the 
best split point only within this variable in a separated second step is a more 
promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
CHAID \citep{Kass:1980}, GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
For the first step of selecting a split variable they all share the same basic concept of 
choosing the covariate which shows the highest association to the response variable 
based on $p$-values provided by a statistical test. While QUEST, CHAID and GUIDE employ 
statistical significance tests for contingency tables, CTree applies permutation tests in 
a conditional inference framework and MOB uses fluctuation tests based on asymptotic theory. 
All these approaches have shown to work well for various situations, however, the specific 
advantages and disadvantages of each of the testing strategies have not yet been 
investigated and compared in detail.

@@

Depending on the setup different effects can be observed for the investigated building blocks.
However, none of the results contradicts the conclusion that in general using all available 
model scores is at least as efficient but mostly clearly advantageous over considering only 
residuals. Also regarding the effects of binarizing residuals or scores at $0$ or categorizing
the values of possible split variables clear recommendations can be made based on the presented 
results.

@@

The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
non-parametric regression tree models in a conditional inference framework by applying 
permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split 
variable~$Z$.
The function $h$ has to depend on $Y$ in a permutation-symmetric way but can be of various forms
and may also be multidimensional. For a numeric response the identity function $h(Y) = Y$ is a common 
choice while a categorical response can be mapped to a unity vector by an indicator function 
$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can return a location and scale parameter of $Y$, 
e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fit to the values of the response variable~$Y$ and regressor 
variable(s)~$X$ the function $h$ can be a model-based transformation returning the 
corresponding model scores. This is done by setting $h(Y) = s(Y,X,\hat{\beta})$
where $\beta$ is the possibly multidimensional model parameter and $s$ is the score function
$$
s(Y, X, \beta) = \frac{\partial \psi(Y, X, \beta)}{\partial \beta}
$$
with the sum of $\psi$ over all observations being the objective function that is optimized 
to obtain the estimated model parameter $\hat{\beta}$. For this purpose many different M-type 
estimators can be applied including maximum likelihood or ordinary least squares such that the 
corresponding objective function can be employed as model-based discrepancy measure for the 
tree algorithm. Evaluationg the score function $s$ on a data set of $n$ observations 
%containing the 
%response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
%regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and on
and a $k$-dimensional estimated parameter vector 
%$\hat{\boldsymbol{\beta}} = (\hat{\beta_1},\ldots,\hat{\beta_k})^\top$ 
yields a $n \times k$-matrix.
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and estimated parameter element 
$(y_i, x_i, \hat{\beta_l})$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$.

@@

To test for independence of $h(Y)$ and $g(Z)$ a linear test statistic is calculated.
Following \cite{Strasser+Weber:1999} the conditional expectation and covariance of the test 
statistic given all permutations of the response variable can be calculated and used to 
standardize the test statistic. In order to map multivariate test statistics into the real 
line two options are given: either the maximum of the absolute values of the standardized 
statistic or a quadratic form. The resulting values can then be transformed to 
the scale of $p$-values where they can be compared in an unbiased way to select a split variable.

If both variables $Y$ and $Z$ are numeric the applied independence test corresponds to a Pearson 
correlation test. For one of the variables being categorical and the other numeric a 1-way ANOVA
is employed while for two categorical variables a $\chi^2$-test is performed. Hence, the CTree algorithm
provides a very general framework where the choice of the type of variables and the corresponding 
functions $g$ and $h$ determine the type of the applied statistical test.
(See \cite{Hothorn+Hornik+VanDeWiel:2006} for examples on how to use the different options of this toolbox.)

@@

\subsection{GUIDE}\label{sec:guide}
Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in 
the GUIDE algorithm \citep{Loh:2002}. Again, simple models such as 
a constant fit whith the parameter $\beta$ being the mean of the response, hence 
$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$ 
%for a response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$, 
can be employed as well as more complex models.
For the statistical test applied to select a split variable some type of model-based
residual is required. In \cite{Loh:2002} linear models are used as an example where the
residuals are defined by
$$
r(Y,X,\hat{\beta}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X.
$$
In related literature other models together with an appropriate choice of residuals
have been applied such as in
regression trees for longitudinal and multiresponse data \citep{Loh+Zheng:2013}, 
%(http://www.stat.rice.edu/~jrojo/4th-Lehmann/slides/Loh.pdf)
quantile regression models \citep{Chaudhuri+Loh:2002} 
and proportional hazards modeling via Poisson regression \citep{Loh+He+Man:2015} among many others.
%Poisson regression trees \citep{Chaudhuri+Lo+Loh:1995}, %(adjusted Anscombe residual (Pierce and Schafer 1986))

@@

In subsequent approaches as in \cite{Hothorn+Hornik+Zeileis:2006}
it is shown that these $p$-values can also be used as a basis for a
pruning strategy namely prepruning: Only if the $p$-value corresponding to the selected split
variable (hence the smallest $p$-value) is smaller than a given significance level a split is
performed while otherwise no further splits are made in the current node.
In that way the information provided by the significance test is exploited in a more profound way.
However, prepruning requires the applied test to perform reasonably well as a significance test, 
i.e. it has to have power but also avoid overfitting.

For postpruning no significance level is included leading to very large trees which are then 
pruned afterwards, for example by employing cost-complexity pruning as in the 
CART algorithm \citep{Breiman+Friedman+Stone:1984}. 
Thus, for postpruning the test only needs to be able to select the best split variable but does 
not have to be effective as a significance test.

Originally the GUIDE algorithm was introduced applying postpruning while CTree and MOB
were both presented employing prepruning (see Table~\ref{tab:combinations} for a more detailed
overview). Alternatively, for model-based partitioning a pruning strategy can also use model-based 
information criteria such as AIC or BIC \citep[see for example][]{Su+Wang+Fan:2004}.
However, all testing strategies can be combined with each of these pruning strategies.
If a test performs well the choice of a pruning strategy has hardly any impact on the results, 
but as prepruning only requires already provided information it is usually less computationally
expensive than the additional calculations needed for postpruning.
However, if the test performs poorly in terms of power and overfitting, prepruning suffers from 
this deficiency while postpruning can compensate it at least partly and improve the results.
For this reason we are focusing on the testing strategies applied in tree algorithms and 
investigate how effective they are as significance tests.

\section{Unifying Framework}\label{sec:unify}
Each of the algorithms CTree, GUIDE, and MOB can be characterized by its combination
of the type of fitted models, the type of applied tests and the employed pruning strategy
as listed in Table~\ref{tab:combinations}. These settings represent the algorithms in their
original form. However, in principle all other combinations of the listed features are possible 
as well. As explained in Section~\ref{sec:pruning} the focus of this paper is on the testing 
strategies (second column of Table~\ref{tab:combinations}). They are here discussed for the case 
of parametric models together with prepruning as the properties of this pruning strategy
are already explained/determined by the testing strategy. Subsequently combinations
with postpruning are evaluated as well to show the impact of the choice of pruning strategy.
From here on the names CTree, MOB, and GUIDE are used to refer to the algorithm-specific
testing strategies rather than the complete algorithms.

@@

\item Categorization of covariates:\\
Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd 
and 3rd quartile are used to bin each possible split variable into four categories. In that way,
depending on whether the scores are binarized or not, a $\chi^2$-test or 1-way ANOVA can be applied
where the quartiles are the only possible split points while otherwise maximally selected splits 
over $Z$ are considered.

@@

\subsection{Linear Model Tree}\label{sec:lmtrees}
As an example a linear model tree is considered as this allows for an easy and well structured 
description of the investigated subjects. However, the conclusions drawn from this example also 
apply for a wide range of other model types. 
The models fitted in the nodes of the linear model tree as considered for the simulation 
study in Section~\ref{sec:simulation} are of the form
$$
Y = \beta_0 + \beta_1 \cdot X + \epsilon
$$
for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
For each node the type of the fitted model is the same such that no node-specific features are 
considered and the observations of the current (arbitrarily chosen) node are denoted by 
$\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.

A very basic form of a discrepancy measure can be obtained for each observation by simply 
calculating deviations of the response values from the mean value within the node, for example 
by $(\bar{y} - y_i)$.
However, this includes only the response variable $Y$ but not the fitted linear model.
For a model-based approach residuals can be considered defined by
$$
r(y_i,x_i,\beta_0,\beta_1) = y_i - \beta_0 - \beta_1 \cdot x_i.
$$
The appropriate choice of the function $\psi$ is then
$$
\psi(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
$$
such that $\sum_{i=1}^n \psi(y_i,x_i,\beta_0,\beta_1)$ is the objective function
which is minimized to obtain the estimated coefficient vector
$\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
The corresponding score function is defined by
$$
s(y_i, x_i, \hat{\beta_0},\hat{\beta_1}) = 
\frac{\partial \psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\beta_0},\hat{\beta_1}) = 
2 \cdot r(y_i,x_i,\hat{\beta_0},\hat{\beta_1}) \cdot (1,x_i)^\top
$$
where the factor $2$ is often omitted such that the first column of the $n \times 2$-dimensional
score matrix corresponds to the residuals. 
Therefore, including only residuals (i.e. only the first column of the score matrix)
corresponds to including the derivatives of $\psi$ only regarding the intercept 
parameter~$\beta_0$.
On the contrary, by using full model scores (i.e. the whole $n \times 2$ matrix) 
the derivatives of $\psi$ regarding the intercept $\beta_0$ (first column) and the slope 
parameter $\beta_1$ (second column) are considered.

@@

As explained before, the aim of this simulation study is to investigate the effects
of each particular building block of the unifying framework presented in Section~\ref{sec:unify} 
rather than the whole testing strategies. Therefore, other combinations as presented in 
Table~\ref{tab:buildingblocks}, hence variations of GUIDE, CTREE, and MOB are evaluated as well
as their original versions. Moreover, as prepruning is employed all of the considered statistical 
tests are applied as significance tests answering two questions at once: 
(1)~Should a split be performed at all? 
(2)~If so, in which variable?
For the first question the $p$-values regarding all available covariates are compared 
to a predefined level of significance $\alpha = 0.05$. Only if the smallest $p$-value is smaller
than $\alpha$ a split is performed and the covariate corresponding to this $p$-value is selected.

@@

Figure~\ref{fig:scores_stump} shows the evaluation results for the three different
scenarios of varying only the intercept $\beta_0$ or only the slope parameter $\beta_1$
or both of them, once for a true split point $\xi=0$ and once for $\xi=0.8$. 
As explained before, the tests are applied as significance tests with significance 
level $\alpha=0.05$. Therefore, the proportion of cases in which the true split variable $Z_1$
is detected with a $p$-value smaller than $\alpha$ is compared.
(To show that in general the drawn conclusions do not change if no significance level is included 
the same setting is evaluated again without any significance level and the results are presented
in Appendix~\ref{app:significance}.)

@@

\subsection{Full Factorial Analysis}\label{sec:3way}
In Figure~\ref{fig:3way} the effects of the building blocks are illustrated for the most general
case where the intercept and slope parameter are both varying and the true split point is in the
center, i.e. $\xi=0$. Based on the results shown in Figure~\ref{fig:scores_stump} the effect size 
$\delta = 0.3$ is considered making the differences particularly well observable. 
(Additional results over increasing $\delta$ as in Section~\ref{sec:scores} 
can be found in Appendix~\ref{app:increasingdelta}.)
The power of the different combinations is measured by the mean $p$-value for the true split 
variable~$Z_1$.
The possible values for each of the three building blocks are listed in Table~\ref{tab:levels}.
In order to evaluate all 12 combinations the original testing strategies CTree, MOB, and GUIDE
are applied together with variations of them. For each combination the corresponding employed
strategy can be found in Appendix~\ref{app:combinations}, Table~\ref{tab:combinations_appendix}.

@@

\newpage
\subsection{Tree}
\label{sec:tree}
To compare the performance of testing strategies in a more complex scenario they have
been evaluated on data sets based on the tree structure with two different split variables 
($Z_1$ and $Z_2$), both with the same split point $\xi$ as illustrated in 
Figure~\ref{fig:dgp_tree}.

@@

\section*{Computational Details}
\label{sec:comp}
The applied implementation is based on the \textsf{R} package \textbf{partykit}
(version~\Sexpr{packageVersion("partykit")}) which is available 
on \textsf{R}-Forge at (\url{https://R-Forge.R-project.org/projects/partykit/}).
The code to reproduce the simulation study can also be found on \textsf{R}-Forge
at (\url{https://R-Forge.R-project.org/projects/partykit/inst/guideliketests}).

>>>>

\section{Introduction}\label{sec:introduction}
In many situations fitting one global model to a given data set can be very challenging, 
especially if the data contains lots of different features with strong variation 
and complex interactions. Therefore, separating the data into more homogeneous subgroups 
based on a set of covariates first can simplify the task and fitting a local model to 
each of the resulting subgroups often leads to better results. This separation can be 
done by applying a tree algorithm. While all of the many developed algorithms 
follow the general idea of splitting the data such that some objective function is optimized, 
they differ in their specific approaches to selecting a split variable and the corresponding 
split point. 
Some of the first tree algorithms (e.g., AID, \citealp{Morgan+Sonquist:1963}; 
CART, \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
the best split point and split variable in one step by directly comparing all possible split
points in all possible split variables. However, it has been shown that this is not only
computationally expensive but also biased towards split variables with many possible split
points \citep{Doyle:1973, Kim+Loh:2001}. 
Therefore, selecting a split variable in a first step and then searching for the 
best split point only within this variable in a separated second step is a more 
promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
CHAID \citep{Kass:1980}, GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
For the first step of selecting a split variable they all share the same basic concept of 
choosing the covariate which shows the highest association to the response variable 
based on $p$-values provided by a statistical test. While QUEST, CHAID and GUIDE employ 
statistical significance tests for contingency tables, CTree applies permutation tests in 
a conditional inference framework and MOB uses fluctuation tests based on asymptotic theory. 
All these approaches have shown to work well for various situations, however, the specific 
advantages and disadvantages of each of the testing strategies have not yet been 
investigated and compared in detail.

@@

Depending on the setup different effects can be observed for the investigated building blocks.
However, none of the results contradicts the conclusion that in general using all available 
model scores is at least as efficient but mostly clearly advantageous over considering only 
residuals. Also regarding the effects of binarizing residuals or scores at $0$ or categorizing
the values of possible split variables clear recommendations can be given based on the presented 
results.

@@

The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
non-parametric regression tree models in a conditional inference framework by applying 
permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split 
variable~$Z$.
The function $h$ has to depend on $Y$ in a permutation-symmetric way but can be of various forms
and may also be multidimensional. For a numeric response the identity function $h(Y) = Y$ is a common 
choice while a categorical response can be mapped to a unity vector by an indicator function 
$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can return a location and scale parameter of $Y$, 
e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fit to the values of the response variable~$Y$ and regressor 
variable(s)~$X$ the function $h$ can be a model-based transformation returning the 
corresponding model scores. This is done by setting $h(Y) = s(Y,X,\hat{\beta})$
where $\beta$ is the possibly multidimensional model parameter and $s$ is the score function
$$
s(Y, X, \beta) = \frac{\partial \psi(Y, X, \beta)}{\partial \beta}
$$
with the sum of $\psi$ over all observations being the objective function that is optimized 
to obtain the estimated model parameter $\hat{\beta}$. For this purpose many different M-type 
estimators can be applied including maximum likelihood or ordinary least squares such that the 
corresponding objective function can be employed as model-based discrepancy measure for the 
tree algorithm. Evaluating the score function $s$ on a data set of $n$ observations 
%containing the 
%response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
%regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and on
and a $k$-dimensional estimated parameter vector 
%$\hat{\boldsymbol{\beta}} = (\hat{\beta_1},\ldots,\hat{\beta_k})^\top$ 
yields a $n \times k$-matrix.
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and estimated parameter element 
$(y_i, x_i, \hat{\beta_l})$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$.

@@

To test for independence of $h(Y)$ and $g(Z)$ a linear test statistic is calculated.
Following \cite{Strasser+Weber:1999} the conditional expectation and covariance of the linear
test statistic given all permutations of the response variable can be calculated and used to 
standardize the test statistic. In order to map multivariate test statistics into the real 
line two options are given: either the maximum of the absolute values of the standardized 
statistic or a quadratic form. The resulting values can then be transformed to 
the scale of $p$-values allowing for an unbiased comparison in order to select a split variable.

If both variables $Y$ and $Z$ are numeric the applied independence test corresponds to a Pearson 
correlation test. For one of the variables being categorical and the other numeric a 1-way ANOVA
is employed while for two categorical variables a $\chi^2$-test is performed. Hence, the CTree 
algorithm provides a very general framework where the type of variables and the corresponding 
functions $g$ and $h$ determine the type of the applied statistical test.
(See \cite{Hothorn+Hornik+VanDeWiel:2006} for examples on how to use the different options of this toolbox.)

@@

\subsection{GUIDE}\label{sec:guide}
Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in 
the GUIDE algorithm \citep{Loh:2002}. Again, simple models such as 
a constant fit with the parameter $\beta$ being the mean of the response, hence 
$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$, 
%for a response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$, 
can be employed as well as more complex models.
For the statistical test applied to select a split variable some type of model-based
residual is required. In \cite{Loh:2002} linear models are used as an example where the
residuals are defined by
$$
r(Y,X,\hat{\beta}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X.
$$
In related literature other models together with an appropriate choice of residuals
have been applied such as in
regression trees for longitudinal and multiresponse data \citep{Loh+Zheng:2013}, 
%(http://www.stat.rice.edu/~jrojo/4th-Lehmann/slides/Loh.pdf)
quantile regression models \citep{Chaudhuri+Loh:2002} 
and proportional hazards modeling via Poisson regression \citep{Loh+He+Man:2015} among many others.
%Poisson regression trees \citep{Chaudhuri+Lo+Loh:1995}, %(adjusted Anscombe residual (Pierce and Schafer 1986))

@@

In subsequent approaches as in \cite{Hothorn+Hornik+Zeileis:2006}
it is shown that these $p$-values can also be used as a basis for a
pruning strategy namely \textit{prepruning}: Only if the $p$-value corresponding to the selected split
variable (hence the smallest $p$-value) is smaller than a given significance level a split is
performed while otherwise no further splits are made in the current node.
In that way the information provided by the significance test is exploited in a more profound way.
However, prepruning requires the applied test to perform reasonably well as a significance test, 
i.e. it has to have power but also avoid overfitting.

On the contrary, when employing postpruning no significance level is included leading to very large 
trees which are then pruned afterwards, for example by applying cost-complexity pruning as in the 
CART algorithm \citep{Breiman+Friedman+Stone:1984}. 
Thus, for postpruning the test only needs to be able to select the best split variable but does 
not have to be effective as a significance test.

Originally the GUIDE algorithm was introduced applying postpruning while CTree and MOB
were both presented employing prepruning (see Table~\ref{tab:combinations} for a more detailed
overview). Alternatively, for model-based partitioning a pruning strategy can also use model-based 
information criteria such as AIC or BIC \citep[see for example][]{Su+Wang+Fan:2004}.
However, all testing strategies can be combined with each of these pruning strategies.
If a test performs well the choice of a pruning strategy has hardly any impact on the results, 
but as prepruning only requires already provided information it is usually less computationally
expensive than postpruning needing additional calculations. 
However, if the test performs poorly in terms of power and overfitting, prepruning suffers from 
this deficiency while postpruning can compensate it at least partly and improve the results.
Therefore, being aware of the power of the selected testing strategy is essential for choosing 
an appropriate pruning strategy.
For that reason we are focusing on the testing strategies applied in tree algorithms 
and investigate how effective they are as significance tests.

\section{Unifying Framework}\label{sec:unify}
Each of the algorithms CTree, GUIDE, and MOB can be characterized by its combination
of the type of fitted models, the type of applied tests and the employed pruning strategy
as listed in Table~\ref{tab:combinations}. These settings represent the algorithms in their
original form. However, in principle all other combinations of the listed features are possible 
as well. As explained in Section~\ref{sec:pruning} the focus of this paper is on the testing 
strategies (second column of Table~\ref{tab:combinations}). They are here discussed for the case 
of parametric models together with prepruning as the properties of this pruning strategy
are already determined by the testing strategy. Subsequently combinations
with postpruning are evaluated as well to show the impact of the choice of pruning strategy.
From here on the names CTree, MOB, and GUIDE are used to refer to the algorithm-specific
testing strategies rather than the complete algorithms.

@@

\item Categorization of covariates:\\
Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd 
and 3rd quartile are considered to bin each possible split variable into four categories. In that way,
depending on whether the scores are binarized or not, a $\chi^2$-test or 1-way ANOVA can be applied
where the quartiles are the only possible split points while otherwise maximally selected splits 
over $Z$ are considered.

@@

\subsection{Linear Model Tree}\label{sec:lmtrees}
As an example a linear model tree is considered as this allows for an easy and well structured 
description of the investigated subjects. However, the conclusions drawn from this example also 
hold for a wide range of other model types. 
The models fitted in the nodes of the linear model tree as considered for the simulation 
study in Section~\ref{sec:simulation} are of the form
$$
Y = \beta_0 + \beta_1 \cdot X + \epsilon
$$
for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
For each node the type of the fitted model is the same such that no node-specific features are 
considered and the observations of the current (arbitrarily chosen) node are denoted by 
$\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.

A very basic form of a discrepancy measure can be obtained for each observation by simply 
calculating deviations of the response values from the mean value within the node, for example 
by $(\bar{y} - y_i)$.
However, this includes only the response variable $Y$ but not the fitted linear model.
For a model-based approach residuals can be considered defined by
$$
r(y_i,x_i,\beta_0,\beta_1) = y_i - \beta_0 - \beta_1 \cdot x_i.
$$
The appropriate choice of the function $\psi$ is then
$$
\psi(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
$$
such that $\sum_{i=1}^n \psi(y_i,x_i,\beta_0,\beta_1)$ is the objective function
which is minimized to obtain the estimated coefficient vector
$\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
The corresponding score function is defined by
$$
s(y_i, x_i, \beta_0, \beta_1) = 
\frac{\partial \psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \beta_0, \beta_1) = 
2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
$$
where the factor $2$ is often omitted such that the first column of the $n \times 2$-dimensional
score matrix corresponds to the residuals. 
Therefore, including only residuals (i.e. only the first column of the score matrix)
corresponds to including the derivatives of $\psi$ only regarding the intercept 
parameter~$\beta_0$.
On the contrary, by using full model scores (i.e. the whole $n \times 2$ matrix) 
the derivatives of $\psi$ regarding the intercept $\beta_0$ (first column) and the slope 
parameter $\beta_1$ (second column) are considered.

@@

As explained before, the aim of this simulation study is to investigate the effects
of each particular building block of the unifying framework presented in Section~\ref{sec:unify} 
rather than the whole testing strategies. Therefore, other combinations as presented in 
Table~\ref{tab:buildingblocks}, hence adapted versions of GUIDE, CTREE, and MOB are evaluated as well
as their original versions. Moreover, as prepruning is employed all of the considered statistical 
tests are applied as significance tests answering two questions at once: 
(1)~Should a split be performed at all? 
(2)~If so, in which variable?
For the first question the $p$-values regarding all available covariates are compared 
to a predefined level of significance $\alpha = 0.05$. Only if the smallest $p$-value is smaller
than $\alpha$ a split is performed and the covariate corresponding to this $p$-value is selected.

@@

Figure~\ref{fig:scores_stump} shows the evaluation results for the three different
scenarios (varying only the intercept $\beta_0$, only the slope parameter $\beta_1$
or both of them) once for a true split point $\xi=0$ and once for $\xi=0.8$. 
As explained before, the tests are applied as significance tests with significance 
level $\alpha=0.05$. Therefore, the proportion of cases in which the true split variable $Z_1$
is detected with a $p$-value smaller than $\alpha$ is compared.
(To show that in general the drawn conclusions do not change if no significance level is included 
the same setting is evaluated again without any significance level and the results are presented
in Appendix~\ref{app:significance}.)

@@

\subsection{Full Factorial Analysis}\label{sec:3way}
In Figure~\ref{fig:3way} the effects of the building blocks are illustrated for the most general
case where the intercept and slope parameter are both varying and the true split point is in the
center, i.e. $\xi=0$. Based on the results shown in Figure~\ref{fig:scores_stump} the effect size 
$\delta = 0.3$ is considered making the differences particularly well observable 
(additional results over increasing $\delta$ as in Section~\ref{sec:scores} 
can be found in Appendix~\ref{app:increasingdelta}).
The power of the different combinations is measured by the mean $p$-value for the true split 
variable~$Z_1$.
The possible levels for each of the three building blocks are listed in Table~\ref{tab:levels}.
In total 12 combinations are applied including the original testing strategies CTree, MOB, and GUIDE
and variations of them. A more detailed overview of all 12 combinations can be found in 
Appendix~\ref{app:combinations}, Table~\ref{tab:combinations_appendix}.

@@

%\newpage
\subsection{Tree}
\label{sec:tree}
To compare the performance of testing strategies in a more complex scenario they have
been evaluated on data sets based on the tree structure with two different split variables 
($Z_1$ and $Z_2$), both with the same split point $\xi$ as illustrated in 
Figure~\ref{fig:dgp_tree}.

@@

\section*{Computational Details}
\label{sec:comp}
The applied implementation is based on the \textsf{R} package \textbf{partykit}
(version~\Sexpr{packageVersion("partykit")}) which is available 
on \textsf{R}-Forge at (\url{https://R-Forge.R-project.org/projects/partykit/})
where the code to reproduce the simulation study can be found as well.
